---
title: "data_exploration.Rmd"
output: html_document
date: "2026-03-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Immigration Data Exploration -- Arrests
Exploring immigration data provided by https://deportationdata.org/

Useful Links:
  - https://deportationdata.org/docs/ice/codebook
  - https://deportationdata.org/data/processed/ice.html 

## init
```{r,echo=FALSE}
library(tidyverse)
library(maps)
library(janitor)
library(here)
library(arrow)
library(readxl)
source("fns.R")
```


## read data


```{r}
arrests_orig <- read_parquet(here("data/march-2026","arrests-latest.parquet"))

```

## high level exploration
```{r}
str(arrests_orig)
```

Basic cleaning
```{r}
arrests <- arrests_orig |> 
  clean_strings() 
```


## uniqueness
```{r}
table(arrests$duplicate_likely)

arrests |>
  filter(duplicate_likely) |>
  count(unique_identifier) |>
  arrange(desc(n))

```

Spot check duplicates
```{r}
arrests |> filter(unique_identifier=="A5A2C5277ED163AAC4C41D7B796FFF7A9793B48E")
```


About 2.5% of rows are flagged as likely duplicates in original dataset, some of 
these are exactly identical (aside from the original ID row). 

### deduplication
1. Remove exact duplicates
2. For those remaining, deduplicate by unique ID and apprehension_date.
  keep those with the most recent departure date
  A. If there are ties, break by which has the most fields populated

Exact duplicates
```{r}
arrests <- arrests |> 
  distinct(pick(-row_original),.keep_all = TRUE)

table(arrests$duplicate_likely)

```
```{r}
# number of non-missing obs
arrests$n_dat <- apply(arrests,1,function(x) sum(!is.na(x)))

# deduplicate
arrests <- arrests |>
  arrange(desc(n_dat)) |>
  distinct(unique_identifier,apprehension_date,.keep_all = TRUE)

```


After deduplication, distinct apprehensions for each individual, also tag the
order/first/most recent apprehensions 

TO DO: make into a function (need to figure out the name injection)
```{r}
arrests <- arrests |>
  arrange(apprehension_date) |>
  mutate(
    n_appr=ifelse(!is.na(unique_identifier), n(), NA),
    appr_ind = ifelse(!is.na(unique_identifier), row_number(), NA),
    first_appr = ifelse(!is.na(unique_identifier), appr_ind==1, NA),
    last_appr = ifelse(!is.na(unique_identifier), appr_ind==n_appr, NA),
    .by=unique_identifier
  )   
  
arrests |> datenum_summary(n_appr)

```

Tag first and most recent arrests
```{r}
arrests <- arrests |>
  arrange(apprehension_date) |>
  mutate(
    appr_ind = row_number(),
    first_appr = appr_ind==1,
    last_appr = appr_ind==n_appr,
    .by=unique_identifier
  ) 

```


## variable checks
### strings

Check missingness
```{r}
arrests |> percent_missing(is.character)

```

Check common values
```{r}
char_vars <- names(select(arrests,is.character))
top_vals <- lapply(char_vars, \(x) arrests |> most_common_vals(x))
names(top_vals) <- char_vars
top_df <- do.call("bind_rows",top_vals)
top_df
```

Questions:
* Which groups are being apprehended in "collateral" apprehensions? How do their
movements across the system compare to others?
* Which countries do people go to most frequently? Which countries often don't 
return to place of origin, which receive those not from their place of origin?
* Is there much variation in arrest method/criminality of arrests by AOR?
* Which states have the most arrests relative to (immigrant) population

### date values
final_order and departed_date have outliers/potential errors. they are also
missing in quite a few cases though I think this is expected b/c they don't 
always occur
```{r}
datenum_summary(arrests,is.Date | is.POSIXct)

```

#### flag potential date errors

I believe that departure date should always be last, though there may be on-the-ground
reasons why this isn't the case.

```{r}
arrests <- arrests |>
  mutate(dep_pre=departed_date < apprehension_date | 
           departed_date<final_order_date) 

arrests |> count(dep_pre)

```

#### flag pre/post
Adding a flag for pre/post Jan 2025 (i.e. the start of Trump's second term).
Starting at Feb 1, which is slightly after the inauguration. A different date
may be more meaningful depending on when policies were enacted on the ground.
```{r}
arrests <- arrests |>
  mutate(post_0125=ifelse(apprehension_date>=as.Date("2025-02-01"),TRUE,FALSE),
    post_0125_f=factor(post_0125, levels=c(TRUE,FALSE),
                        labels=c("After Jan 2025","Before Jan 2025")
    )
#    post_0125_lab=ifelse(post_0125,"")
  ) 
  
summary(arrests$post_0125)
```

#### order of/time between events

```{r}
arrests <- arrests |>
  mutate(
    ord_appr = apprehension_date-final_order_date,    
    appr_dep = departed_date-apprehension_date,
    ord_dep = departed_date-final_order_date
  )

arrests |>
  summarise(
    ord_appr_mean = mean(ord_appr,na.rm=TRUE),
    ord_appr_median = median(ord_appr,na.rm=TRUE),
    appr_dep_mean = mean(appr_dep,na.rm=TRUE),
    appr_dep_median = median(appr_dep,na.rm=TRUE),
    ord_dep_mean = mean(ord_dep,na.rm=TRUE),
    ord_dep_median = median(ord_dep,na.rm=TRUE)
  ) |> 
  t()

```

Checking for missingness of apprehension/departure date. 

First, missingness that seems due to data issues
```{r}
arrests |> 
  filter(final_order_yes_no=="YES" & is.na(final_order_date))
arrests |> 
  filter(final_order_yes_no=="NO" & !is.na(final_order_date))

arrests |> 
  filter(is.na(departure_country) & !is.na(departed_date))
arrests |> 
  filter(!is.na(departure_country) & is.na(departed_date))


```

Generally, when a departure/final order variable indicates that the event should
happen there is also a date for that event (and vice versa). The most common is
missing departure country with departure date, but very few of those. For now
just ignoring all of these cases.

Now looking at other cases when apprehension/departure date is missing (assuming
there's an administrative reason for the most part).


Checking order of events by combinations of both any two and all three steps.

```{r}
arrests <- arrests |> 
  mutate(
    ord_appr_proc = case_when(
      ord_appr > 0 ~ "ord-appr",
      ord_appr < 0 ~ "appr-ord",
      ord_appr == 0 ~ "same",
      TRUE ~ NA
    ),
    appr_dep_proc = case_when(
      appr_dep > 0 ~ "appr-dep",
      appr_dep < 0 ~ "dep-appr",
      appr_dep == 0 ~ "same",
      TRUE ~ NA
    ),
    ord_dep_proc = case_when(
      ord_dep > 0 ~ "ord-dep",
      ord_dep < 0 ~ "dep-ord",
      ord_dep == 0 ~ "same",
      TRUE ~ NA
    )
  )

arrests |> count(ord_appr_proc)
arrests |> count(appr_dep_proc)
arrests |> count(ord_dep_proc)
```


```{r}
# three steps
arrests <- arrests |> 
  mutate(
    proc_action3 = case_when(
      is.na(apprehension_date) | 
        is.na(final_order_date) | 
        is.na(departed_date) ~ "<3 steps",
      apprehension_date < final_order_date &
        final_order_date < departed_date ~ "appr-ord-dep",
      final_order_date < apprehension_date & 
        apprehension_date < departed_date ~ "ord-appr-dep",
      final_order_date < departed_date &  
        departed_date < apprehension_date ~ "ord-dep-appr",
      apprehension_date < departed_date & 
      departed_date < final_order_date ~ "appr-dep-ord",
      departed_date < final_order_date & 
      final_order_date < apprehension_date ~ "dep-ord-appr",
      departed_date < apprehension_date & 
      apprehension_date < final_order_date ~ "dep-appr-ord",
      apprehension_date < departed_date & 
      departed_date == final_order_date ~ "appr-ordDep",
      apprehension_date < departed_date & 
      apprehension_date == final_order_date ~ "apprOrd-dep",
      final_order_date < departed_date & 
      apprehension_date == departed_date ~ "ord-apprDep",      
      apprehension_date == departed_date &
        departed_date < final_order_date ~ "apprDep-ord", 
      apprehension_date == final_order_date &
        departed_date < final_order_date ~ "dep-apprOrd", 
      departed_date == final_order_date &
        departed_date < apprehension_date ~ "depOrd-appr", 
      apprehension_date == departed_date &
        departed_date == final_order_date ~ "all-same", 
      TRUE ~ "other"
    )
  )

arrests |> 
  count(proc_action3) |>
  arrange(desc(n))

```


How do these trends look based on ICE group, criminality, program, etc?
How have those trends changed over time?

```{r}
arrests |> 
  filter(apprehension_date < as.Date("2025-01-20")) |>
  count(ord_appr_proc,apprehension_criminality) |>
  pivot_wider(names_from=apprehension_criminality,values_from = n)


arrests |> 
  filter(apprehension_date >= as.Date("2025-01-20")) |>
  count(ord_appr_proc,apprehension_criminality) |>
  pivot_wider(names_from=apprehension_criminality,values_from = n)


```

From a quick check, it looks like post-Trump2 (just using 01/20/2025 as the cutoff),
there are more cases of individuals without criminal charges who have a deportation
order and then were arrested compared to pre-Trump2


#### Apprehension trends by date and time
Get day of week and hour of apprehension
```{r}
arrests <- arrests |>
  mutate(
    apprehension_dow=wday(apprehension_date_time,label=TRUE),
    apprehension_hour=hour(apprehension_date_time)
  )


```

```{r}
  arrests |> 
  filter(apprehension_type=="COLLATERAL") |>
  filter(apprehension_date>=as.Date("2025-01-01")) |>
  count(apprehension_dow,apprehension_hour,apprehension_type) |>
  ggplot(aes(x=apprehension_dow,y=factor(apprehension_hour))) +
  geom_tile(aes(fill=n)) 

  arrests |> 
  filter(apprehension_type=="TARGETED") |>
  filter(apprehension_date>=as.Date("2025-01-01")) |>
  count(apprehension_dow,apprehension_hour,apprehension_type) |>
  ggplot(aes(x=apprehension_dow,y=factor(apprehension_hour))) +
  geom_tile(aes(fill=n)) 

```
Looks like collateral apprehensions occur at a slightly broader range of days/times
while targeted apprehensions are generally during "business hours," with a pretty
strong peak between 9 and 3 Tuesdays through Fridays


### logical 

```{r}
logical_summary(arrests)
```

Only logical in original is from the duplicate likely flag, which is nearly 
always populated (and rarely true)

### integer
Create age variables from YOB
```{r}
arrests |> datenum_summary(is.numeric)
```

Only numerics from original data are birth_year and row_number. from an initial 
check, birth_year seems to have a reasonable range and very low missingness.

Create a variable for age at apprehension (note that this won't be exact since
we don't have full birthday).

```{r}
arrests <- arrests |>
  mutate(appr_age=year(apprehension_date)-birth_year)

arrests |> select(birth_year,apprehension_date,appr_age) |> head()

```


## demographics

Main demo variables:
  * citizenship_country
  * apprehension_state
  * gender
  * appr_age

### age and gender
Create age bins
```{r}
# Consider making this function run from 
arrests <- make_age_bins(arrests,col=appr_age)

table(arrests$age_bin)
arrests |> 
  group_by(age_bin) |>
  summarise(min(appr_age),max(appr_age),n())

```


Chart age and gender over time
```{r}
arrests |>
  mutate(appr_year=year(apprehension_date)) |>
  ggplot(aes(x=age_bin,fill=gender)) +
  geom_bar() +
  ggplot2::facet_grid(rows=vars(appr_year),scales="free_y")
```
Apprehensions have been of older groups, and more males in recent years. 
Could this be because there is less border enforcement and more enforcement
at workplaces, etc? Could potentially look at apprehension_method and 
final_program to further disaggregate.


Cross-tabs for age and gender
```{r}
arrests |> 
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents()


arrests |> 
  filter(apprehension_criminality=="3 OTHER IMMIGRATION VIOLATOR") |>
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents()


```

### geographies
Create regions based on apprehension_state and country variables

TO DO: Make these functions

#### states
state: base on census regions? or perhaps the modal apprehension_aor
  look into location caveats on documentation

```{r}
state_aor_groups <- arrests |> 
  group_by(apprehension_state,apprehension_aor) |>
  summarise(n=n()) |> 
  arrange(desc(n)) |>
  mutate(ind = row_number()) |>
  filter(ind==1) |>
  ungroup() |>
  arrange(apprehension_state)


```

Map results (may want to combine all AORs from the same states, probably 
depends on case)
```{r}
state_aor_groups <- state_aor_groups |>
  mutate(region=str_to_lower(apprehension_state))

map_data("state") |> 
  left_join(state_aor_groups) |>
  arrange(order) |>
  ggplot(aes(long, lat)) +
    geom_polygon(aes(group = group, fill = apprehension_aor)) +
    coord_map("albers",  lat0 = 45.5, lat1 = 29.5)
```

Add aor back to arrest data
```{r}
state_aor_groups <- state_aor_groups |> 
  rename(aor_group=apprehension_aor) |>
  select(apprehension_state, aor_group)

arrests <- arrests |>
  left_join(state_aor_groups) |>
  relocate(aor_group,.after=apprehension_aor)

```

```{r}
arrests |>
  count(aor_group,apprehension_state)
```


#### countries
UN Geogregions from https://unstats.un.org/unsd/methodology/m49/overview/ 
NOTE: for now only keeping region, but some other categories could be
interesting disaggregates (developing country statuses)
```{r}
un_geo_orig <- read_delim(here("data/other-data","UNSDMethodology.csv"),
                     delim=";") |>
  janitor::clean_names()
un_geo <- un_geo_orig |>
  mutate(country=str_to_upper(country_or_area)) |>
  mutate(region=str_to_upper(
    coalesce(intermediate_region_name,sub_region_name))) |>
  select(country,region)

un_geo 

# manual matches 
country_bridge <- read_csv(here("data/other-data",
                                  "ICE_UN_country_manual_matches.csv")) 

```

Clean country names
```{r}
clean_country_names <- function(str) {
  str |> 
    str_remove_all(",.*$") |>
    str_replace_all("[^[:alnum:]]", " ") |>
    str_replace_all("  "," ") |>
    str_trim()
}

un_geo <- un_geo |> 
  rename(country_orig=country) |>
  mutate(country=clean_country_names(country_orig),.after=country_orig) |>
  left_join(country_bridge,by=(c("country"="un_country"))) |>
  mutate(country=ifelse(!is.na(citizenship_country),citizenship_country,
                        country)) |>
  distinct(country,.keep_all = TRUE) |>
  select(country,region)

```



```{r}
if(!"citizenship_country_orig" %in% names(arrests)){
  arrests <- arrests |> 
    rename(citizenship_country_orig=citizenship_country) |>
    mutate(citizenship_country=clean_country_names(citizenship_country_orig),
                                                   .after=citizenship_country_orig) |>
    rename(departure_country_orig=departure_country) |>
    mutate(departure_country=clean_country_names(departure_country_orig),
                                                   .after=departure_country_orig) 
  
  
}
```


Join to arrests data
```{r}
arrests <- arrests |>
  left_join(un_geo,by=join_by(citizenship_country==country)) |>
  rename(citizenship_region=region) |>
  relocate(citizenship_region,.after=citizenship_country) |> 
  left_join(un_geo,by=join_by(departure_country==country)) |>
  rename(departure_region=region) |>
  relocate(departure_region,.after=departure_country)

```

Check missings for name differences

(quite a bit of remapping to do, could potentially remove special characters and
words after commas)

NOTE: Make sure there are no duplicated names (.x and .y) 
```{r}
arrests |>
  filter(is.na(citizenship_region) & !is.na(citizenship_country)) |>
  distinct(citizenship_country) |>
  arrange(citizenship_country) 
```

```{r}
arrests |> 
  count(citizenship_region,departure_region) |> 
  arrange(desc(n))
#arrests |> count(citizenship_region) |> arrange(desc(n))
```

For highly-represented countries (>10,000?), set to the country rather than region
```{r}
arrests |>
  count(citizenship_country) |>
  arrange(desc(n))

use_countries <- arrests |>
  count(citizenship_country) |>
  filter(n>10000) |>
  select(citizenship_country) |>
  unlist()

```


```{r}
arrests <- arrests |>
  mutate(citizenship_region=ifelse(citizenship_country %in% use_countries,
                                   citizenship_country,citizenship_region)) |>
  mutate(departure_region=ifelse(departure_country %in% use_countries,
                                   departure_country,departure_region))  
```


Which countries do people go to most frequently? Which countries often don't 
return to place of origin, which receive those not from their place of origin?

```{r}

arrests |>
  filter(!is.na(departure_country)) |>
  filter(departed_date >= as.Date("2025-02-01")) |> 
  add_count(citizenship_country,name="n_from") |>
  add_count(departure_country,name="n_to") |>  
  filter(n_from>1500 | n_to>1500) |>
  filter(departure_country!=citizenship_country) |> 
  group_by(citizenship_country,departure_country) |>
  summarise(n=n()) |>
  filter(n>25) |> 
  ggplot(aes(departure_country, citizenship_country)) +
    geom_tile(aes(fill = n)) +
    scale_fill_gradient(low = "pink", high = "darkred")  +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```


Same analysis based on world region

```{r}
arrests |>
  filter(!is.na(departure_country)) |>
  filter(departed_date >= as.Date("2025-02-01")) |> 
  add_count(citizenship_region,name="n_from") |>
  add_count(departure_region,name="n_to") |>  
  #filter(n_from>1500 | n_to>1500) |>
  filter(departure_region!=citizenship_region) |> 
  group_by(citizenship_region,departure_region) |>
  summarise(n=n()) |>
  filter(n>25) |> 
  ggplot(aes(departure_region, citizenship_region)) +
    geom_tile(aes(fill = n)) +
    scale_fill_gradient(low = "pink", high = "darkred")  +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```



```{r}
arrests |> 
#  filter(str_detect(departure_country,"CHINA")) |>
filter(citizenship_country=="CANADA" & departure_country!= "CANADA")

# 5-year old canadian sent to romania? 
  # 	326C3B206EF5C8160F3E964B3597ADDC725FD270

# check others sent to romania on that day
# TODO: check within a month or two as well as the exact date
arrests |> 
  filter(departure_country=="ROMANIA" & apprehension_date==as.Date("2025-09-11"))
```


## Export
Save cleaned file with addl columns 
```{r}
write_parquet(arrests,here("data/cleaned","arrests.parquet"))

```

