---
title: "explore_detainers_data.Rmd"
output: html_document
date: "2026-03-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Immigration Data Exploration -- Detainers
Exploring immigration data provided by https://deportationdata.org/

Useful Links:
  - https://deportationdata.org/docs/ice/codebook
  - https://deportationdata.org/data/processed/ice.html 

## init
```{r,echo=FALSE}
library(tidyverse)
library(maps)
library(janitor)
library(here)
library(arrow)
library(readxl)
library(sf)
library(sfarrow)
source("fns.R")
```


## read data


```{r}
list.files(path=here("data/march-2026"),"parquet")
```

```{r}
detention_orig <- read_parquet(here("data/march-2026",
                                    "detention-stays-latest.parquet"))

```


### facility locations
```{r}
facilities_orig <- st_read_parquet(here("data/other-data",
                                    "facilities-latest-sf.parquet"))
```


## high level exploration
```{r}
str(detention_orig)
```

```{r}
#View(facilities_orig)
```


Detentions data is very large, so using a sample for exploration
```{r}
detention_orig <- detention_orig |> 
  slice_sample(n=200000)
```


Basic cleaning
```{r}
detention <- detention_orig |> 
  janitor::clean_names() |>
  clean_strings() 

facilities <- facilities_orig |> 
  janitor::clean_names() |>
  clean_strings()

```


For most uses of the facilities data here, not all of the columns are needed.
```{r}
facilities <- facilities |>
  select(detention_facility_code,name,address,city,county,county_fips_code,
         core_based_statistical_area,combined_statistical_area,state,zip)
```

Create a version of facilities without geometry
```{r}
facilities_nogeo <- facilities |> st_drop_geometry()
```



## uniqueness


### deduplication
TO DO: Check this more closely
1. Remove exact duplicates
2. For those remaining, deduplicate by unique id and detainer_prepare_date
  A. If there are ties, break by which has the most fields populated

Exact duplicates (don't seem to be any)
```{r}
detention <- detention |> 
  distinct(pick(-stay_id),.keep_all = TRUE) 

```

```{r}
detention <- detention |> 
  group_by(stay_id) |>
  mutate(n_obs=n())
```


```{r}
#detainers |>
#  filter(!duplicate_likely & n_obs > 1) |>
#  arrange(unique_identifier,detainer_prepare_date) |>
#  select(unique_identifier,detainer_prepare_date,detainer_type,detention_facility_code,
#         birth_year,citizenship_country,gender,msc_charge_date,msc_charge_code,
#         case_status,case_category,final_order_date,departed_date,illegal_reentry_yes_no,
#         resume_custody_yes_no,detainer_lift_reason)
```



```{r}
# number of non-missing obs
#detention$n_dat <- apply(detention,1,function(x) sum(!is.na(x)))

# deduplicate
#detention <- detention |>
#  arrange(desc(detention_prepare_date),desc(n_dat)) |>
#  distinct(stay_id,detainer_prepare_date,.keep_all = TRUE)

```


Stats on detention counts
```{r}
detention <- detention |>
  ungroup() |>
  arrange(book_in_date_time_first) |>
  mutate(
    n_det=ifelse(!is.na(unique_identifier), n(), NA),
    det_ind = ifelse(!is.na(unique_identifier), row_number(), NA),
    first_det = ifelse(!is.na(unique_identifier), det_ind==1, NA),
    last_det = ifelse(!is.na(unique_identifier), det_ind==n_det, NA),
    .by=unique_identifier
  )   
  
detention |> datenum_summary(n_det)

```

### facilities
```{r}
facilities_nogeo |> 
  count(detention_facility_code) |>
  filter(n>1)
```


## variable checks
### strings

Check missingness
```{r}
detention |> percent_missing(is.character)

```

Check common values
```{r}
char_vars <- names(select(detention,is.character))
top_vals <- lapply(char_vars, \(x) detention |> most_common_vals(x))
names(top_vals) <- char_vars
top_df <- do.call("bind_rows",top_vals)
top_df
```

Notes:
1. Lots of variables with criminality information

### date values
Key date seems to be `book_in_date_time`,


```{r}
datenum_summary(detention,is.Date | is.POSIXct)

```

This doesn't have apprehension date. It would be interesting to see if there are
any gaps.

#### flag potential date errors
TO DO

### logical 

```{r}
logical_summary(detention)
```

All logicals are from self-generated columns

### numeric

```{r}
detention |> datenum_summary(is.numeric) |> round(1)
```

Very high amount of missingness, it may not actually be plausible to do much with these.
There are also several variables with a max of repeated 9s. It seems like these may indicate missingness (or some sort of data error). 

Checking frequency
```{r}
detention |> filter(initial_bond_set_amount==999999) |> nrow()
detention |> filter(msc_sentence_years==999) |> nrow()
detention |> filter(msc_sentence_months==999) |> nrow()
```

Based on how infrequent they are, going to set them to NA.
```{r}
detention <- detention |> mutate(
  initial_bond_set_amount = replace_values(initial_bond_set_amount, 999999~NA),
  msc_sentence_years = replace_values(msc_sentence_years, 999~NA),  
  msc_sentence_months = replace_values(msc_sentence_months, 999~NA)
)
```

```{r}
detention |> datenum_summary(is.numeric) |> round(1)
```

Potential questions (though missingness makes this harder)
What factors lead to longer sentences? 
Particularly, what factors beyond criminality?
Would need to look at coding MSC to do this

Create a variable for age at detention (note that this won't be exact since
we don't have full birthday).

```{r}
detention <- detention |>
  mutate(det_age=year(book_in_date_time_first)-birth_year)


```


## demographics

Main demo variables:
  * citizenship_country/birth_country
  * gender
  * race
  * det_age
  * location of first detention

### age and gender
Create age bins
```{r}
detention <- make_age_bins(detention,col=det_age)

detention |> 
  group_by(age_bin) |>
  summarise(min(det_age),max(det_age),n())

```


Chart age and gender over time
```{r}
detention |>
  mutate(det_year=year(book_in_date_time_first)) |>
  filter(det_year>=2022) |>
  ggplot(aes(x=age_bin,fill=gender)) +
  geom_bar() +
  ggplot2::facet_grid(rows=vars(det_year),scales="free_y")
```
As with all datasets, detentions are largely male. As with arrests they have 
become less common for females in recent years (opposite trend with detainers).



Cross-tabs for age and gender
```{r}
detention |> 
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents()


detention |> 
  filter(book_in_criminality=="3 OTHER IMMIGRATION VIOLATOR") |>
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents()


```

### race
Set Unknown race to NA
```{r}
detention <- detention |>
  mutate(race = replace_values(race, "UNKNOWN"~NA))
```

```{r}
detention |>
  count(race,ethnicity) |>
  arrange(desc(n))
```



### geographies
NOTE: Is initial detention a reasonable proxy, or should i look at book_in_site
Think data documentation has some info on this

Join facilities data to get location of initial detention

```{r}
facilities <- facilities |>
  rename(cbsa=core_based_statistical_area) |>
  rename(csa=combined_statistical_area) |>
  rename_with(~paste0("detention_first_",.x),-detention_facility_code)
```

```{r}
detention <- detention |>
  left_join(facilities,by=c("detention_facility_code_first"="detention_facility_code"))
```


```{r}
detention |>
  filter(str_detect(detention_first_cbsa,"MO$|KS$")) |>
  count(detention_first_cbsa)
```
```{r}
detention |>
#  filter(str_detect(detention_first_csa,"MO$|KS$")) |>
  count(detention_first_csa)
```

```{r}
facilities |>
  filter(str_detect(detention_first_cbsa,"MO$|KS$")) |>
  count(detention_first_cbsa)
```


## conditions of detention
TO DO
Levels of Criminality, case status/category,


## Export
Save cleaned files  
```{r}
write_parquet(facilities,here("data/cleaned","facilities_sf.parquet"))
write_parquet(facilities_nogeo,here("data/cleaned","facilities_list.parquet"))

```


