---
title: "explore_detainers_data.Rmd"
output: html_document
date: "2026-03-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Immigration Data Exploration -- Detainers
Exploring immigration data provided by https://deportationdata.org/

Useful Links:
  - https://deportationdata.org/docs/ice/codebook
  - https://deportationdata.org/data/processed/ice.html 

## init
```{r,echo=FALSE}
library(tidyverse)
library(maps)
library(janitor)
library(here)
library(arrow)
library(readxl)
source("fns.R")
```


## read data

```{r}
list.files(path=here("data/march-2026"),"parquet")
```


```{r}
detainers_orig <- read_parquet(here("data/march-2026","detainers-latest.parquet"))

```

## high level exploration
```{r}
str(detainers_orig)
```

Basic cleaning
```{r}
detainers <- detainers_orig |> 
  clean_strings() 


```


## uniqueness
```{r}
table(detainers$duplicate_likely)

detainers |>
  filter(duplicate_likely) |>
  count(unique_identifier) |>
  arrange(desc(n))

```

Spot check duplicates
```{r}
detainers |> filter(unique_identifier=="108F71C5B85009BA9550C4B79ADEFD4D6390EE69")
```


About 4.1% of rows are flagged as likely duplicates in original dataset, some of 
these are exactly identical (aside from the original ID row). 

### deduplication
1. Remove exact duplicates
2. For those remaining, deduplicate by unique id and detainer_prepare_date
  A. If there are ties, break by which has the most fields populated

Exact duplicates
```{r}
detainers <- detainers |> 
  distinct(pick(-row_original),.keep_all = TRUE) 

table(detainers$duplicate_likely)

```

```{r}
detainers <- detainers |> 
  group_by(unique_identifier) |>
  mutate(n_obs=n())
```


```{r}
detainers |>
  filter(!duplicate_likely & n_obs > 1) |>
  arrange(unique_identifier,detainer_prepare_date) |>
  select(unique_identifier,detainer_prepare_date,detainer_type,detention_facility_code,
         birth_year,citizenship_country,gender,msc_charge_date,msc_charge_code,
         case_status,case_category,final_order_date,departed_date,illegal_reentry_yes_no,
         resume_custody_yes_no,detainer_lift_reason)
```



```{r}
# number of non-missing obs
detainers$n_dat <- apply(detainers,1,function(x) sum(!is.na(x)))

# deduplicate
detainers <- detainers |>
  arrange(desc(detainer_prepare_date),desc(n_dat)) |>
  distinct(unique_identifier,detainer_prepare_date,.keep_all = TRUE)

```



```{r}
detainers <- detainers |>
  ungroup() |>
  arrange(detainer_prepare_date) |>
  mutate(
    n_det=ifelse(!is.na(unique_identifier), n(), NA),
    det_ind = ifelse(!is.na(unique_identifier), row_number(), NA),
    first_det = ifelse(!is.na(unique_identifier), det_ind==1, NA),
    last_det = ifelse(!is.na(unique_identifier), det_ind==n_det, NA),
    .by=unique_identifier
  )   
  
detainers |> datenum_summary(n_det)

```



## variable checks
### strings

Check missingness
```{r}
detainers |> percent_missing(is.character)

```

Check common values
```{r}
char_vars <- names(select(detainers,is.character))
top_vals <- lapply(char_vars, \(x) detainers |> most_common_vals(x))
names(top_vals) <- char_vars
top_df <- do.call("bind_rows",top_vals)
top_df
```

Questions:
1. What are the criminal statuses of those with detainers?
2. What factors are most likely to lead to a detainer?

Look at detention sites in KS and MO
```{r}
detainers |>
  filter(facility_state %in% c("MISSOURI","KANSAS")) |>
  count(facility_state,detention_facility) |>
  arrange(desc(n))

```
Quite a few in KS/MO area. Browsing the first few pages there is: 
Wyandotte, Johnson, KCPD/KCK, Jackson,Independence, Clay, Belton

What percent of KC arrests lead to detainers? What characteristics do they have? 


### date values
Key date seems to be `detainer_prepare_date` which is when the detainer (extension of jail stay) was issued. It is never missing, though there are some in the future (presumably
a data issue but not sure).

Date overview (in order I _think_ they would tend to go in, though am sure there 
is a lot of variation):
* entry_date
* apprehension_date
* msc_charge_date
* msc_conviction_date 
* detainer_prepare_date
* final_order_date
* departed_date


```{r}
datenum_summary(detainers,is.Date | is.POSIXct)

```

#### flag potential date errors
Detainer date in the future
```{r}
detainers |>
  filter(detainer_prepare_date >= as.Date("2026-03-11"))
```

Very few records from after date of data

*TO DO* Order of events




### logical 

```{r}
logical_summary(detainers)
```

Only logical in original is from the duplicate likely flag, worth noting that
it's missing in 24% of cases

### numeric


```{r}
detainers |> datenum_summary(is.numeric)
```

Might be more meaningful from detentions, but what factors lead to longer
sentences? Particularly, what factors beyond criminality?
Would need to look at coding MSC to do this

Create a variable for age at apprehension (note that this won't be exact since
we don't have full birthday).

```{r}
detainers <- detainers |>
  mutate(det_age=year(detainer_prepare_date)-birth_year)

detainers |> select(birth_year,detainer_prepare_date,det_age) |> head()

```


## demographics

Main demo variables:
  * citizenship_country
  * apprehension_state
  * gender
  * det_age

### age and gender
Create age bins
```{r}
detainers <- make_age_bins(detainers,col=det_age)

detainers |> 
  group_by(age_bin) |>
  summarise(min(det_age),max(det_age),n())

```


Chart age and gender over time
```{r}
detainers |>
  mutate(det_year=year(detainer_prepare_date)) |>
  ggplot(aes(x=age_bin,fill=gender)) +
  geom_bar() +
  ggplot2::facet_grid(rows=vars(det_year),scales="free_y")
```
Detainers are largely male, but have become more common for females in recent
years (opposite trend with arrests)


Cross-tabs for age and gender
```{r}
detainers |> 
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents()


detainers |> 
  filter(detainer_prepared_criminality=="3 OTHER IMMIGRATION VIOLATOR") |>
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents()


```

### geographies
TO DO
