---
title: "Get KC Apprehensions"
output: html_document
date: "June 2026"
author: "Dylan Knaggs"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

# Identifying Apprehensions in Kansas City

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Set up environment
```{r,echo=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(arrow)
library(gt)
source("fns.R")
```


## Get Data
```{r}
arrests <- read_parquet(here("data/cleaned","arrests.parquet"))
```

## Initial Location Checks
Per data documentation from the Immigration Data Project and review, the main indicators 
of apprehension location seem to be `toa_current_duty_site` and 
`apprehension_site_landmark`. Checking values of each.

### `toa_current_duty_site`
```{r}
arrests |>
  count(toa_current_duty_site)

```

### `apprehension_site_landmark`
```{r}
arrests |>
  count(apprehension_site_landmark)
``` 

There are a huge number of values. Checking the more common ones.

```{r}
arrests |>
  count(apprehension_site_landmark) |>
  arrange(desc(n)) |>
  head(1000) |>
  arrange(apprehension_site_landmark)

```


While it is a new variable (as of March 2026 release) without documentation, 
`toa_current_duty_site` seems much cleaner, with only a few values that seem to
indicate KC. Using as an initial proxy.
```{r}
arrests <- arrests |>
  mutate(kc_duty_site=str_detect(toa_current_duty_site,"KANSAS") & 
           str_detect(toa_current_duty_site,"MO"))
```


## Statewide Checks

### KC `current_duty_site`, outside MO/KS 

Starting by looking at all KS and MO arrests (based on `apprehension_state`), and 
checking the agreement between state and `current_duty_site`.

```{r}
# Total count
sum(arrests$kc_duty_site & 
      !(arrests$apprehension_state=="MISSOURI" | 
          arrests$apprehension_state == "KANSAS"),na.rm=TRUE)

# View cases
arrests |> 
  filter(kc_duty_site &
           !(arrests$apprehension_state=="MISSOURI" | 
               arrests$apprehension_state == "KANSAS")) |>
  select(toa_current_duty_site,apprehension_site_landmark,
         apprehension_state,apprehension_date)
```

Very few disagreements and almost all of them seem like they are out of
the region. For now going to start with KS and MO as a prerequisite for 
being considered a KC arrest.

### Other `current_duty_site` in MO/KS
Also looking at values of `current_duty_site` in MO/KS that aren't flagged. 
```{r}
arrests |> 
  filter(!kc_duty_site &
           (arrests$apprehension_state=="MISSOURI" | 
               arrests$apprehension_state == "KANSAS")) |>
  count(apprehension_state,toa_current_duty_site,apprehension_site_landmark)
```

There are a handful of duty sites outside of either state, but generally
they all seem to be in non-KC areas of the states.

### Flag State
Based on the above, filtering to only include KS and MO arrests.

```{r}
arrests_moks <- arrests |> 
  filter(apprehension_state=="MISSOURI" | apprehension_state == "KANSAS")

```



## Variable-Specific Checks
With this filter in place, checking values in each location indicator column
```{r}
arrests_moks |> count(apprehension_site_landmark) |> arrange(desc(n))
arrests_moks |> count(toa_current_duty_site) |> arrange(desc(n))

```

It looks like the initial `current_duty_site` flag captured relevant cases but it
is now easier to see relevant `apprehension_site_landmark`.

### Apprehension Site
Note that this is based on manual data inspection and may need to be updated
for future data releases.
```{r}
metro_keywords <- c("CLAY","JACKSON","DOUGLAS","LEAVENWORTH","GRANDVIEW",
                    "INDEPENDENCE","JACKSON","JOHNSON","KANSAS CITY",
                    "PLATTE","WYANDOTTE","CASS")
arrests_moks <- arrests_moks |> 
  mutate(kc_landmark=str_detect(apprehension_site_landmark,
                                paste0(metro_keywords,collapse="|")) &
  !(str_detect(apprehension_site_landmark,"JOHNSON") & 
    apprehension_state=="MISSOURI") &
  !(str_detect(apprehension_site_landmark,"JACKSON") & 
           apprehension_state=="KANSAS")
  )
  


```


### Compare results
Look at overlap/differences between the two indicators 
```{r}
paste("KC by landmark but not duty site:",
sum(arrests_moks$kc_landmark & !arrests_moks$kc_duty_site,na.rm=TRUE)
)

paste("KC by duty site but not landmark:",
sum(!arrests_moks$kc_landmark & arrests_moks$kc_duty_site,na.rm = TRUE)
)

paste("KC by both:",
sum(arrests_moks$kc_landmark & arrests_moks$kc_duty_site,na.rm = TRUE)
)

paste("KC by neither:",
sum(!arrests_moks$kc_landmark & !arrests_moks$kc_duty_site,na.rm = TRUE)
)
```

#### KC Duty Site but not KC Landmark
There are lots of values for duty site but not landmark in KC. Reviewing these.

All results
```{r}
arrests_moks |> 
  filter(kc_duty_site & !kc_landmark) |>
  select(toa_current_duty_site,apprehension_site_landmark,
         apprehension_state,apprehension_date,apprehension_method) 

```


Unique values of each variable
```{r}
# duty site
arrests_moks |> 
  filter(kc_duty_site & !kc_landmark) |>
  count(toa_current_duty_site)

# landmark
arrests_moks |> 
  filter(kc_duty_site & !kc_landmark) |>
  count(apprehension_site_landmark) |>
  arrange(desc(n))


```

The majority of the cases with initial disagreement have an `apprehension_site_landmark`
of "KAN GENERAL AREA, NON-SPECIFIC". It's hard to say for sure, but seems plausible that
these are in the KCMO area. It is noteworthy though that a notable number of records 
are clearly marked with an `apprehension_site_landmark` in other areas

#### KC Landmark but not KC Duty Site
Also checking values for landmark in kc but duty site elsewhere (this is much rarer).

All results
```{r}
arrests_moks |> 
  filter(!kc_duty_site & kc_landmark) |>
  select(toa_current_duty_site,apprehension_site_landmark,
         apprehension_state,apprehension_date,apprehension_method) 
```

Unique values of each variable
```{r}
# duty site
arrests_moks |> 
  filter(!kc_duty_site & kc_landmark) |>
  count(toa_current_duty_site)

# landmark
arrests_moks |> 
  filter(!kc_duty_site & kc_landmark) |>
  count(apprehension_site_landmark) |>
  arrange(desc(n))

```

All of these seem to be outside of the KC metro and can probably be safely 
excluded.

## Final Assignment of KC Arrests

Final logic:
Consider as Kansas City if the `current_duty_site` indicates KCMO, unless the
`apprehension_site_landmark` clearly indicates it is elsewhere. 

For `apprehension_site_landmark` values of "KAN GENERAL AREA, NON-SPECIFIC", 
consider as KCMO but add an `likely` flag. Unfortunately this does leave over
half of the records as only "likely" so additional approaches should be taken to
more clearly identify KC records.

```{r}
arrests_kc <- arrests_moks |>
  filter(
    (kc_duty_site & kc_landmark) |  
      (kc_duty_site & 
         (apprehension_site_landmark=="KAN GENERAL AREA, NON-SPECIFIC" |
            is.na(apprehension_site_landmark)
          )
       )
    ) |>
  mutate(likely=(kc_duty_site & 
    (apprehension_site_landmark=="KAN GENERAL AREA, NON-SPECIFIC" |
       is.na(apprehension_site_landmark)
      )
  ))
```

```{r}
paste("Total arrests flagged in KC:",nrow(arrests_kc))

paste("Records with potential unclarity:",sum(arrests_kc$likely))
```


### Check `likely` flag
Is there a trend in the increase in lack of certainty?
```{r}
arrests_kc <- arrests_kc |>
  mutate(apprehension_month = floor_date(apprehension_date,unit="months")) 

arrests_kc |>
  ggplot(aes(x=apprehension_month,color=likely)) +
  geom_line(stat="count")

```

The likely flag increases greatly post-Jan 25. Based on the fact that other data
seems to be of lower quality in this time period, my suspicion is that this is 
largely down to decreased data quality, leading to more generic values in
`apprehension_site_landmark`.

If this is the case, the KC dataset still may be an overestimate, but not
necessarily dramatically so.


### (not in use) Ever arrested in KC
The below also adds arrests of indviduals who have ever been apprehended in KC.
Not currently using it as it only adds a handful and is outside of focus at the moment.
```{r}
#kc_inds <- arrests_kc$unique_identifier
#arrests_kc_ever <- arrests |>
#  filter(unique_identifier %in% kc_inds)
```

### Export 
```{r}
write_parquet(arrests_kc,here("data/cleaned","arrests_kc.parquet"))
```

