---
title: "explore_detainers_data.Rmd"
output: html_document
date: "2026-03-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Immigration Detentions in KC Area
Exploring immigration data provided by https://deportationdata.org/

Useful Links:
  - https://deportationdata.org/docs/ice/codebook
  - https://deportationdata.org/data/processed/ice.html 

## init
```{r,echo=FALSE}
library(tidyverse)
library(maps)
library(janitor)
library(here)
library(arrow)
library(readxl)
library(sf)
library(sfarrow)
source("fns.R")
```


## read data

### detentions
```{r}
detention_orig <- read_parquet(here("data/march-2026",
                                    "detention-stays-latest.parquet"))

```


### facility locations
```{r}
facilities_orig <- st_read_parquet(here("data/other-data",
                                    "facilities-latest-sf.parquet"))
```



Detentions data is very large, so using a sample for exploration
```{r}
detention_orig <- detention_orig |> 
  slice_sample(n=500000)
```


Basic cleaning
```{r}
detention <- detention_orig |> 
  janitor::clean_names() |>
  clean_strings() 

facilities <- facilities_orig |> 
  janitor::clean_names() |>
  clean_strings()

```


For most uses of the facilities data here, not all of the columns are needed.
```{r}
facilities <- facilities |>
  select(detention_facility_code,name,address,city,county,county_fips_code,
         core_based_statistical_area,combined_statistical_area,state,zip)
```

Create a version of facilities without geometry
```{r}
facilities_nogeo <- facilities |> st_drop_geometry()
```


## Get site geography
NOTE: Is initial detention a reasonable proxy, or should i look at book_in_site
Think data documentation has some info on this

Join facilities data to get location of initial detention

```{r}
facilities <- facilities |>
  rename(cbsa=core_based_statistical_area) |>
  rename(csa=combined_statistical_area) |>
  rename_with(~paste0("detention_first_",.x),-detention_facility_code)
```

```{r}
detention <- detention |>
  left_join(facilities,by=c("detention_facility_code_first"="detention_facility_code"))
```


```{r}
detention |>
  filter(str_detect(detention_first_cbsa,"MO$|KS$")) |>
  count(detention_first_cbsa)
```
```{r}
detention |>
#  filter(str_detect(detention_first_csa,"MO$|KS$")) |>
  count(detention_first_csa)
```

```{r}
facilities |>
  filter(str_detect(detention_first_cbsa,"MO$|KS$")) |>
  count(detention_first_cbsa)
```



## Export
Save cleaned file with addl columns 

