---
title: "KC Analysis"
output: html_document
date: "2026-04-04"
---

# Analysis of ICE data in Kansas City


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

High-level Plan
1. Start with apprehension data and filter down based on toa_curent_duty_site
  should compare with other indicators
2. Left join on detention data
3. Look at time/distance from apprehension to first detention location
4. Start with detention data, and see who is booked into KC area facilities
5. Check the overlap of this group with #2 
(e.g. those with a first stay in KC but an apprehension elsewhere and vice versa)
  See if documentation discusses why this could happen
6. Replicate #3
7. Perform other analyses flagged in scripts

TODO: For Arrests, look at toa_current_duty_site (think this is a new column in March 26 data)
TODO: Look at individuals in one dataset but not others, especially if it seems
like they should be (eg everyone should have an apprehension)
TODO: Keep ppl who have ever been apprehended in KC in addition to apprehensions 
that occur in KC


## Set up environment
```{r,echo=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(arrow)
source("fns.R")
```


## Get Data

```{r}
arrests <- read_parquet(here("data/cleaned","arrests.parquet"))

```

Adding a flag for pre/post Jan 2025
```{r}
arrests <- arrests |>
  mutate(post_0125=apprehension_date>=as.Date("2025-01-01"))
  
```


## Filter for Kansas City Data

Check if there could be KC arrests with the state marked (incorrectly) as neither
KS or MO

```{r}
arrests <- arrests |>
  mutate(kc_duty_site=str_detect(toa_current_duty_site,"KANSAS") & 
           str_detect(toa_current_duty_site,"MO"))

```

```{r}
sum(arrests$kc_duty_site & 
      !(arrests$apprehension_state=="MISSOURI" | 
          arrests$apprehension_state == "KANSAS"),na.rm=TRUE)

arrests |> 
  filter(kc_duty_site &
           !(arrests$apprehension_state=="MISSOURI" | 
               arrests$apprehension_state == "KANSAS")) |>
  select(toa_current_duty_site,apprehension_site_landmark,
         apprehension_state,apprehension_date)
```

Very few potential diagreements and almost all of them seem like they are out of
the region. For now going to start with KS and MO as a prerequisite for 
being considered a KC arrest.


Checking via both `apprehension_site_landmark` and `toa_current_duty_site`. The latter
seems much cleaner but it is also a new variable without much documentation.

Start by looking at only KS and MO state

```{r}
arrests_moks <- arrests |> 
  filter(apprehension_state=="MISSOURI" | apprehension_state == "KANSAS")

```


Checking values in each column
```{r}

arrests_moks |> count(apprehension_site_landmark) |> arrange(desc(n))
arrests_moks |> count(toa_current_duty_site) |> arrange(desc(n))

```

### `apprehension_site_landmark`
Tagging specific mentions of KS and MO counties in the KC metro within `apprehension_site_landmark`
```{r}
metro_keywords <- c("CLAY","JACKSON","DOUGLAS","LEAVENWORTH","GRANDVIEW",
                    "INDEPENDENCE","JACKSON","JOHNSON","KANSAS CITY",
                    "PLATTE","WYANDOTTE","CASS")
arrests_moks <- arrests_moks |> 
  mutate(kc_landmark=str_detect(apprehension_site_landmark,
                                paste0(metro_keywords,collapse="|")) &
  !(str_detect(apprehension_site_landmark,"JOHNSON") & 
    apprehension_state=="MISSOURI") &
  !(str_detect(apprehension_site_landmark,"JACKSON") & 
           apprehension_state=="KANSAS")
  )
  


```


### `toa_current_duty_site`
```{r}
arrests_moks <- arrests_moks |>
  mutate(kc_duty_site=str_detect(toa_current_duty_site,"KANSAS") & 
           str_detect(toa_current_duty_site,"MO"))
```


### Compare results
Overlap/differences 
```{r}
# KC by landmark but not duty site 
sum(arrests_moks$kc_landmark & !arrests_moks$kc_duty_site,na.rm=TRUE)

# KC by duty site but not landmark
sum(!arrests_moks$kc_landmark & arrests_moks$kc_duty_site,na.rm = TRUE)

# KC by both
sum(arrests_moks$kc_landmark & arrests_moks$kc_duty_site,na.rm = TRUE)

# KC by neither
sum(!arrests_moks$kc_landmark & !arrests_moks$kc_duty_site,na.rm = TRUE)

```

Values for duty site in kc but landmark not
```{r}
arrests_moks |> 
  filter(kc_duty_site & !kc_landmark) |>
  select(toa_current_duty_site,apprehension_site_landmark,
         apprehension_state,apprehension_date,apprehension_method) 

arrests_moks |> 
  filter(kc_duty_site & !kc_landmark) |>
  count(toa_current_duty_site)

arrests_moks |> 
  filter(kc_duty_site & !kc_landmark) |>
  count(apprehension_site_landmark) |>
  arrange(desc(n))



```

If there is a KC `toa_current_duty_site`, going to flag as potentially in KC if 
`apprehension_site_landmark` is "KAN GENERAL AREA, NON-SPECIFIC" and otherwise
definitely not in KC.


Values for landmark in kc but duty site not
```{r}
arrests_moks |> 
  filter(!kc_duty_site & kc_landmark) |>
  select(toa_current_duty_site,apprehension_site_landmark,
         apprehension_state,apprehension_date,apprehension_method) 

arrests_moks |> 
  filter(!kc_duty_site & kc_landmark) |>
  count(toa_current_duty_site)

arrests_moks |> 
  filter(!kc_duty_site & kc_landmark) |>
  count(apprehension_site_landmark) |>
  arrange(desc(n))



```

All of these seem to be outside of the KC metro

### Final Assignment of KC Arrests
```{r}
arrests_kc <- arrests_moks |>
  filter(
    (kc_duty_site & kc_landmark) |  
  (kc_duty_site & 
     (
       apprehension_site_landmark=="KAN GENERAL AREA, NON-SPECIFIC" |
         is.na(apprehension_site_landmark)
      )
  )) |>
  mutate(unclear= (kc_duty_site & 
     (
       apprehension_site_landmark=="KAN GENERAL AREA, NON-SPECIFIC" |
         is.na(apprehension_site_landmark)
      )
  ))
```

Also going to make a dataset of arrests of indviduals who have ever been apprehended in kc
(currently this only adds a handful).

```{r}
#kc_inds <- arrests_kc$unique_identifier
#arrests_kc_ever <- arrests |>
#  filter(unique_identifier %in% kc_inds)

```




# Analysis 

## Number of arrests over time
```{r}
arrests_kc <- arrests_kc |>
  mutate(apprehension_month = floor_date(apprehension_date,unit="months")) 

arrests_kc |>
  ggplot(aes(x=apprehension_month)) +
  geom_line(stat="count")


```


## Apprehension Types
```{r}
arrests_kc |> 
  count(apprehension_method) |> 
  arrange(desc(n))

```


Trying to group apprehension methods
```{r}
arrests_kc <- arrests_kc |>
  mutate(
    apprehension_summ = factor(
      case_when(
        apprehension_method %in% c("LOCATED","NON-CUSTODIAL ARREST") ~ 
          "Community",
        str_detect(apprehension_method,"^CAP .* INCARCERATION$") ~ 
        "Prison/Jail",
        apprehension_method == "CUSTODIAL ARREST" ~ 
          "Prison/Jail",      
        TRUE ~ "Others"
      ), levels = c("Community","Prison/Jail","Others")
    )
  )
  

arrests_kc |>
  count(apprehension_summ)

```


```{r}
arrests_kc |>
  ggplot(aes(x=apprehension_month,fill=apprehension_summ)) +
  geom_bar(stat="count")
```

## Criminality
```{r}
arrests_kc |>
  ggplot(aes(x=apprehension_month,fill=apprehension_criminality)) +
  geom_bar(stat="count")

```

```{r}
arrests_kc |>
  count(post_0125,apprehension_criminality) |>
  pivot_wider(names_from=post_0125,values_from=n) |>
  column_percents()

```


## Trends by day of week and time of day
The below looks only at arrests within communities.

```{r}
arrests_kc <- arrests_kc |>
  mutate(
    apprehension_dow=wday(apprehension_date_time,label=TRUE),
    apprehension_hour=hour(apprehension_date_time),
    apprehension_block=cut(
      apprehension_hour,
      breaks=6,
#      breaks=age_bins,
#      labels=age_labs,
      right=TRUE,
      include.lowest=TRUE,
      dig.lab=0,
      ordered_result=TRUE
    )
  )


```


TO DO: Make these nice labels
```{r}
  arrests |>
  mutate(
    apprehension_hour=hour(apprehension_date_time),
    apprehension_block=cut(
      apprehension_hour,
      breaks=12,
      right=TRUE,
      include.lowest=TRUE,
      dig.lab=0,
      ordered_result=TRUE
    )
  ) |>
  filter(apprehension_method %in% c("NON-CUSTODIAL ARREST","LOCATED")) |>
  filter(apprehension_date>=as.Date("2025-01-01")) |>
  count(apprehension_dow,apprehension_block) |>
  ggplot(aes(x=apprehension_dow,y=apprehension_block)) +
  geom_tile(aes(fill=n)) 


```


## By Demographics

### Place of Citizenship
Overall totals
```{r}
arrests_kc |>
  count(citizenship_region) |>
  arrange(desc(n))


```



Individuals with a different `departure_country` vs `citizenship_country`
```{r}
arrests_kc |>
  filter(!is.na(departure_country) & departure_country!=citizenship_country) |>
  filter(!post_0125) |>
  count(citizenship_country,departure_country) |>
  arrange(desc(n))

arrests_kc |>
  filter(!is.na(departure_country) & departure_country!=citizenship_country) |>
  filter(post_0125) |>
  count(citizenship_country,departure_country) |>
  arrange(desc(n))

```

### Age and Gender
```{r}
arrests_kc |>
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents()
```

```{r}
arrests_kc |>
  count(age_bin,post_0125) |>
  pivot_wider(names_from=post_0125,values_from=n) |>
  column_percents() 

```


```{r}
arrests_kc |>
  count(gender,post_0125) |>
  pivot_wider(names_from=post_0125,values_from=n) |>
  column_percents() 

```

TO DO: Make a fn for comparing variables pre and post 012025. 
  should be like the above with a change column added
TO DO: Update label for post-012025

# outcomes (in comparison to country?)


