---
title: "KC Analysis-Overview"
output: html_document
date: "2026-04-04"
---

TO DO: Run everything with the likely flag

# High Level Analysis of ICE data in Kansas City

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

NOTE: mark as preliminary due to lack of definitiveness in current duty site


High-level Plan
1. Start with apprehension data and filter down based on toa_curent_duty_site
  should compare with other indicators
2. Left join on detention data
3. Look at time/distance from apprehension to first detention location
4. Start with detention data, and see who is booked into KC area facilities
5. Check the overlap of this group with #2 
(e.g. those with a first stay in KC but an apprehension elsewhere and vice versa)
  See if documentation discusses why this could happen
6. Replicate #3
7. Perform other analyses flagged in scripts


TODO:
1. split out KC identification notebook
Consider:
2. create notebook for summary (demographics, timeline, apprehension types, day and hr trends)
3. Create notebook for criminality
4. Create notebook for tactics (apprehension types, day and hr trends)

## Set up environment
```{r,echo=FALSE}
library(tidyverse)
library(ggrepel)
library(here)
library(arrow)
source("fns.R")
```

Repeated plot elements
```{r}
theme_set(theme_bw())
inaug_line <- geom_vline(xintercept = as.Date("2025-01-20"), 
                         color="darkblue", size=.75)

inaug_text <- annotate("text",label="Trump's Second Inauguration",
#           x=as.Date("2025-06-28"),y=10,size=3) + # bottom right
           x=as.Date("2024-08-01"),y=180,size=3)  # top left
```


## Get Data

```{r}
arrests_kc <- read_parquet(here("data/cleaned","arrests_kc.parquet"))

```


# Analysis 

## Overview

### Total Arrests

```{r}
# removing partial month of March 2026
arrests_kc |>
  filter(apprehension_month!=as.Date("2026-03-01")) |>
  ggplot(aes(x=apprehension_month)) +
  inaug_line + 
  inaug_text + 
#  geom_vline(xintercept = as.Date("2025-01-20"), color="darkblue", size=.75) +
#  annotate("text",label="Trump's Second Inauguration",
#           x=as.Date("2025-06-28"),y=10,size=3) + # bottom right
#           x=as.Date("2024-08-01"),y=180,size=3) + # top left
  geom_line(stat="count") +
  geom_point(stat="count") +
  geom_text_repel(aes(label = after_stat(count)), 
                  stat = "count", hjust = 0.5,size=2.75)+ #,
#                  linewidth=0,fill="grey80",alpha=.5) + # code for geom_label
  labs(title="Number of ICE Arrests in the Kansas City Area",
       x="Month",y="Number of Apprehensions") +
  scale_x_date(date_breaks = "6 months", date_labels =  "%b %y") +
  ylim(0,NA)

```
### Demographics

#### Place of Citizenship
Apprehensions from citizens of most countries have increased at least somewhat 
since Jan 2025, though there are some exceptions (e.g., Nicaragua and Cuba).
Countries with the notably large increases include Honduras and Venezuela.

I don't know enough about changing immigration patterns and policies to know the
exact causes of these changes.

```{r}
arrests_kc |> 
  comp_table(citizenship_region) |>
  make_output_cols() |>
  add_before_after_change()

```

#### People Deported to Countries of which They Aren't Citizens   
For arrested individuals who are later deported, this table shows how many
individuals were sent to different country than the one for which they have
citizenship.

This is quite rare for Kansas City arrests, though it has increased since Jan 
2025, with Mexico being the most common departure country by far.
```{r}
arrests_kc |>
  filter(!is.na(departure_country) & departure_country!=citizenship_country) |>
  filter(!post_0125) |>
  count(citizenship_country,departure_country,name="number_of_people") |>
  arrange(desc(number_of_people)) |>
  make_output_cols()

arrests_kc |>
  filter(!is.na(departure_country) & departure_country!=citizenship_country) |>
  filter(post_0125) |>
  count(citizenship_country,departure_country,name="number_of_people") |>
  arrange(desc(number_of_people)) |>
  make_output_cols()

```

#### Percent of Arrests by Age and Gender

Each column in this table shows the percent each age group makes up of all
arrests within a gender. For example, 6-10 year olds make up 4.4% of all female
arrests.

For both men and women, people in their 20s and 30s make up over half of all
arrests.
```{r}
arrests_kc |>
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents() |>
  make_output_cols() |>
  select(-`Unknown`)
```


This table shows changes in the number of arrests by age. Arrests of people under 
18 have (maybe surprisingly) decreased since Jan 2025. However, 18-21 year olds
are the age group with the largest increase. 
```{r}
arrests_kc |>
  comp_table(age_bin,sort_by_var = TRUE) |>
  make_output_cols() |>
  add_before_after_change() 

```

Looking at the number of arrests by gender shows that arrests of men have driven
the increase since Jan 2025.
```{r}
arrests_kc |>
  comp_table(gender,sort_by_var = TRUE) |>
  make_output_cols() |>
  add_before_after_change() 

```

## Apprehension Tactics
### Places of Apprehensions
Looking at all values of the `apprehension_method` column.
```{r}
arrests_kc |> 
  count(apprehension_method) |> 
  arrange(desc(n))

```


Trying to group apprehension methods to get place information
```{r}
arrests_kc <- arrests_kc |>
  mutate(
    apprehension_summ = factor(
      case_when(
        apprehension_method %in% c("LOCATED","NON-CUSTODIAL ARREST") ~ 
          "Community",
        str_detect(apprehension_method,"^CAP .* INCARCERATION$") ~ 
        "Prison/Jail",
        apprehension_method == "CUSTODIAL ARREST" ~ 
          "Prison/Jail",      
        TRUE ~ "Others"
      ), levels = c("Community","Prison/Jail","Others")
    )
  )
  

arrests_kc |>
  count(apprehension_summ)

```


```{r}
arrests_kc |>
  filter(apprehension_date!=as.Date("2026-02-01")) |>
  ggplot(aes(x=apprehension_month,fill=apprehension_summ)) +
  geom_bar(stat="count") +
  inaug_line + 
  inaug_text 
  



```

### Apprehension Programs

```{r}
arrests_kc |>
  count(post_0125_f,final_program) |>
  arrange(desc(post_0125_f)) |>
  pivot_wider(names_from=post_0125_f,values_from = n) |>
  column_percents()
```


### Days with the most apprehensions
```{r}
arrests_kc |>
  count(apprehension_date) |>
  arrange(desc(n))
```


```{r}
arrests_kc |>
  filter(apprehension_date == as.Date("2025-06-03") | 
           apprehension_date == as.Date("2025-06-04"))
```


### Trends by day of week and time of day

Focusing on arrests within communities.

```{r}

hr_bins <- seq(0,23,2)
hr_labs <- c(
  "12 AM - 2 AM",
  "2 AM - 4 AM",
  "4 AM - 6 AM",
  "6 AM - 8 AM",
  "8 AM - 10 AM",
  "10 AM - 12 PM",
  "12 PM - 2 PM",
  "2 PM - 4 PM",
  "4 PM - 6 PM",
  "6 PM - 8 PM",
  "10 PM - 12 AM"
  )
arrests_kc <- arrests_kc |>
  mutate(
    apprehension_dow=wday(apprehension_date_time,label=TRUE),
    apprehension_hour=hour(apprehension_date_time),
    apprehension_block=cut(
      apprehension_hour,
      breaks=hr_bins,
      labels=hr_labs,
      right=TRUE,
      include.lowest=TRUE,
      dig.lab=0,
      ordered_result=TRUE
    )
  )


```

TO DO:
Title
Theme (remove axis lines and maybe use something else)
Note (caveat about not a prediction)
Wrap legend title

```{r}
  arrests_kc |>
  filter(apprehension_method %in% c("NON-CUSTODIAL ARREST","LOCATED")) |>
  filter(post_0125) |>
  count(post_0125,apprehension_dow,apprehension_block) |>
  ggplot(aes(x=apprehension_dow,y=apprehension_block)) +
    geom_tile(aes(fill=n)) +
    scale_fill_continuous(low="grey95",high="red") +
    labs(x="Day of Week",y="Hour",fill="Number of Apprehensions") 
  


```




## Criminality

### Overall criminality
Trend of arrests on people convicted of a crime
```{r}
arrests_kc |>
  ggplot(aes(x=apprehension_month,fill=apprehension_criminality)) +
  geom_bar(stat="count") 


```

### Severity 
For those convicted of a crime, severity of the crime by Case Threat Level. Per the Deportation Data Project,
potentially based on criteria from a [2011 ICE Memo](https://www.ice.gov/doclib/news/releases/2011/110302washingtondc.pdf)

* Level 1: Aggravated felony
* Level 2: Felony or at least 3 misdemeanors
* Level 3: Misdemeanors

```{r}
arrests_kc |> count(case_threat_level)
```


Combining criminality and severity 
```{r}
crim_levels <- 5:0
crim_labels <- c(
  "No Criminal Charges",
  "Criminal Charges, No Convictions",
  "Conviction: 1-2 Misdemeanors",
  "Conviction: Felony/3 or More Misdemeanors",    
  "Conviction: Aggravated Felony",
  "Conviction: Unknown Severity"
)


arrests_kc <- arrests_kc |> 
  mutate(crim_summ=
    case_when(
      is.na(apprehension_criminality) ~ NA,
      apprehension_criminality == "3 OTHER IMMIGRATION VIOLATOR" ~ 5,
      apprehension_criminality == "2 PENDING CRIMINAL CHARGES" ~ 4,
      case_threat_level==3 ~ 3,
      case_threat_level==2 ~ 2,    
      case_threat_level==1 ~ 1,
      TRUE ~ 0
    )
  )

arrests_kc <- arrests_kc |>
  mutate(crim_summ=factor(crim_summ,levels = crim_levels,labels=crim_labels))

```




```{r}
arrests_kc |>
  filter(!is.na(crim_summ)) |>
  ggplot(aes(x=apprehension_month,fill=crim_summ)) +
  geom_bar(stat="count")

arrests_kc |>
  filter(!is.na(crim_summ)) |>
  ggplot(aes(x=apprehension_month,fill=crim_summ)) +
  geom_bar(stat="count")

```



### Changes in current administration

Criminality
```{r}
arrests_kc |>
  count(post_0125,apprehension_criminality) |>
  pivot_wider(names_from=post_0125,values_from=n) |>
  column_percents()

```


Threat level
```{r}

crim_levels <- 5:0
crim_labels <- c(
  "No Criminal Charges",
  "Criminal Charges, No Convictions",
  "Conviction: 1-2 Misdemeanors",
  "Conviction: Felony/3 or More Misdemeanors",    
  "Conviction: Aggravated Felony",
  "Conviction: Unknown Severity"
)

arrests_kc |> 
  mutate(crim_summ=
    case_when(
      is.na(apprehension_criminality) ~ NA,
      apprehension_criminality == "3 OTHER IMMIGRATION VIOLATOR" ~ 5,
      apprehension_criminality == "2 PENDING CRIMINAL CHARGES" ~ 4,
      case_threat_level==3 ~ 3,
      case_threat_level==2 ~ 2,    
      case_threat_level==1 ~ 1,
      TRUE ~ 0
    )
  ) |>
  mutate(crim_summ=factor(crim_summ,levels = crim_levels,labels=crim_labels)) |>
  mutate(apprehension_month = floor_date(apprehension_date,unit="months")) |>
  filter(!is.na(case_threat_level)) |>
  count(post_0125,apprehension_month,crim_summ) |>
  group_by(post_0125,crim_summ) |>
  summarise(n_avg=mean(n)) |>
  pivot_wider(names_from=post_0125,values_from=n_avg) |>
  mutate(
    change=`TRUE`-`FALSE`,
    pct_change=100*(`TRUE`-`FALSE`)/`FALSE`
  )

arrests_kc |> 
  filter(!likely) |>
  mutate(crim_summ=
    case_when(
      is.na(apprehension_criminality) ~ NA,
      apprehension_criminality == "3 OTHER IMMIGRATION VIOLATOR" ~ 5,
      apprehension_criminality == "2 PENDING CRIMINAL CHARGES" ~ 4,
      case_threat_level==3 ~ 3,
      case_threat_level==2 ~ 2,    
      case_threat_level==1 ~ 1,
      TRUE ~ 0
    )
  ) |>
  mutate(crim_summ=factor(crim_summ,levels = crim_levels,labels=crim_labels)) |>
  mutate(apprehension_month = floor_date(apprehension_date,unit="months")) |>
  filter(!is.na(case_threat_level)) |>
  count(post_0125,apprehension_month,crim_summ) |>
  group_by(post_0125,crim_summ) |>
  summarise(n_avg=mean(n)) |>
  pivot_wider(names_from=post_0125,values_from=n_avg) |>
  mutate(
    change=`TRUE`-`FALSE`,
    pct_change=100*(`TRUE`-`FALSE`)/`FALSE`
  )
  



arrests_kc |> 
  mutate(crim_summ=
    case_when(
      is.na(apprehension_criminality) ~ NA,
      apprehension_criminality == "3 OTHER IMMIGRATION VIOLATOR" ~ 5,
      apprehension_criminality == "2 PENDING CRIMINAL CHARGES" ~ 4,
      case_threat_level==3 ~ 3,
      case_threat_level==2 ~ 2,    
      case_threat_level==1 ~ 1,
      TRUE ~ 0
    )
  ) |>
  mutate(crim_summ=factor(crim_summ,levels = crim_levels,labels=crim_labels)) |>
  mutate(apprehension_month = floor_date(apprehension_date,unit="months")) |>
  filter(!is.na(case_threat_level)) |>
  count(post_0125_f,apprehension_month,crim_summ) |>
  group_by(post_0125_F,crim_summ) |>
  summarise(n_avg=mean(n)) |>
  pivot_wider(names_from=post_0125,values_from=n_avg) |>
  mutate(
    change=`TRUE`-`FALSE`,
    pct_change=100*(`TRUE`-`FALSE`)/`FALSE`
  )
  
  

```

NOTE: huge increase in Criminal Charges, No Convictions could have something to do with the way
KC cases are identified


### Case Statuses
```{r}
arrests_kc |> 
  count(case_status) 
```

```{r}
arrests_kc |>
  count(case_threat_level)
```



