---
title: "Kansas City ICE Apprehension Analysis"
date: "June 2026"
author: "Dylan Knaggs"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide  
---

The below document provides some quantitative insights into Immigration 
and Customs Enforcement (ICE) activity within the Kansas City metro area between
October 2022 and February 2026, particularly focusing on the changes seen during
the current presidential administration.

Some key takeaways:

* The overall rate of arrests has more than tripled during the current administration,
with particularly high numbers during the summer of 2025 and February 2026.
* Eighteen to twenty-one year-olds have experienced the largest increase in ICE arrests,
with a 278% increase.
* During the current administration, there have been particular peaks in ICE activity
within communities on Tuesday and Wednesday from 8-10 AM.
  - Note that this does not be mean that this will continue to be the case in the future.
* Arrests do not seem targeted towards those with the most severe criminal background. 
  - People convicted of aggravated felonies have seen the smallest increase in arrests.
  - The groups with the largest increases are people convicted of 1-2 misdemeanors
  and those with a criminal charge but no convictions. 

I would love to collaborate both with additional analysts and those who are 
working on immigration issues. For questions and feedback, feel free to reach out at [dylan.knaggs@gmail.com](mailto:dylan.knaggs@gmail.com).

All analysis is based on data provided in response to a FOIA request to the [Deportation Data Project](https://deportationdata.org/). Many thanks to them for making this data 
and for making their work available to the public.

# Configuration
```{r,collapse=TRUE,results='hide',message=FALSE,warning=FALSE}
library(tidyverse)
library(knitr)
library(gt)
library(ggrepel)
library(here)
library(arrow)
source("fns.R")
```

Repeated plot elements
```{r,collapse=TRUE,results='hide',message=FALSE,warning=FALSE}
theme_set(theme_bw())
inaug_line <- geom_vline(xintercept = as.Date("2025-01-20"), 
                         color="darkblue", size=.75)

add_inaug_text <- function(text_date=as.Date("2024-08-01")){
  annotate("text",label="Trump's Second Inauguration",
           x=text_date,y=180,size=3)  # top left
#           x=as.Date("2025-06-28"),y=10,size=3) + # bottom right
  
}

color_vec <- c("Before Jan 2025"="#E69A8D","After Jan 2025"="#5F4B8B")
loc_color_vec <- c("Community"="salmon3","Prison/Jail"="palegreen4",
                   "Others"="grey75")

```


## Load Data
For simplicity in plotting and calculations, removing March 2026, which only
has data for a few days.
```{r,collapse=TRUE,results='hide',message=FALSE,warning=FALSE}
arrests_kc <- read_parquet(here("data/cleaned","arrests_kc.parquet")) |>
    filter(apprehension_month!=as.Date("2026-03-01")) 
```

## Calculate repeated values
```{r,collapse=TRUE,results='hide',message=FALSE,warning=FALSE}
key_dates <- arrests_kc |> 
  summarise(min=min(apprehension_date),max=max(apprehension_date))
key_dates_admin <- arrests_kc |>
  filter(post_0125) |>
  summarise(min=min(apprehension_date),max=max(apprehension_date))
first_date <- key_dates$min
first_date_admin <- key_dates_admin$min
last_date <- key_dates_admin$max
```


# Analysis 

## Overview

### Total Arrests

#### Before and after Trump II Inauguration

The average number of arrests each month within the Kansas City area has more 
than tripled from Oct 2022-Jan 2025 to 
Feb 2025-Feb 2026.

```{r,collapse=TRUE}
arrests_kc |>
  mutate(n_arr=1) |>
  comp_table(n_arr) |>
  add_before_after_change() |>
  select(-n_arr) |>
  gt_defaults("Average Monthly Arrests",decimals=1)
```


#### Number of arrests over time
```{r,collapse=TRUE}
arrests_kc |>
  ggplot(aes(x=apprehension_month)) +
  inaug_line + 
  add_inaug_text() + 
  geom_line(stat="count") +
  geom_point(stat="count") +
  geom_text_repel(aes(label = after_stat(count)), 
                  stat = "count", hjust = 0.5,size=2.75)+ #,
#                  linewidth=0,fill="grey80",alpha=.5) + # code for geom_label
  labs(title="Number of ICE Arrests in the Kansas City Area",
       x="Month",y="Number of Apprehensions") +
  scale_x_date(date_breaks = "6 months", date_labels =  "%b %y") +
  ylim(0,NA)

```

### Demographics

#### Place of citizenship

Citizens of Mexico have been apprehended more than any other country both before and after the second Trump inauguration. However, apprehensions from citizens of countries throughout the world have increased  since Jan 2025, though countries with the notably large increases 
include Honduras, Guatemala, Venezuela, and El Salvador.


```{r,collapse=TRUE}
country_tbl <- arrests_kc |> 
  comp_table(citizenship_region,month_norm=TRUE) |>
  make_output_cols() 

country_tbl |>
  add_before_after_change() |>
  head(15) |>
  gt_defaults("Average Monthly Arrests by Country/Region of Citizenship") 
```

```{r}
n <- 10
country_tbl |>
  head(10) |>
  pivot_for_plot() |>
  ggplot(aes(x=reorder(`Citizenship Region`,`value`,decreasing=FALSE),y=`value`,
             label=signif(`value`,2),fill=`Period`)) +
  geom_col(position="dodge",width=.7) +
  scale_fill_manual(values=color_vec) +
  geom_text(position = position_dodge(width = .7),vjust=-.3,size=2.75) +   
  labs(
    title="Average Monthly Arrests by Country of Citizenship",
    subtitle=paste0(
      "Kansas City Area, ",
      format(first_date,"%b %Y"),"-",
      format(last_date,"%b %Y"),"\n",
      "Top ",n, " Countries with the Most Apprehensions"
    ),
    x="Citizenship Country",y="Average Monthly Arrests") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```



##### Deportations to third countries  
For arrested individuals who are later deported, this table shows how many
individuals were sent to a different country than the one for which they have
citizenship.

This is fairly uncommon for Kansas City arrests, though it has increased since Jan 
2025, with Mexico being the most common departure country by far.
```{r,collapse=TRUE}
arrests_kc |>
  mutate(
    departure_country=str_to_title(departure_country),
    citizenship_country=str_to_title(citizenship_country)
  ) |>
  filter(!is.na(departure_country) & departure_country!=citizenship_country) |>
  filter(!post_0125) |>
  count(citizenship_country,departure_country,name="number_of_people") |>
  arrange(desc(number_of_people)) |>
  make_output_cols() |>
  gt() |>
  tab_header("Arrests Leading to Departures to Third Countries (Oct 2022-Jan 2025)")
```

```{r,collapse=TRUE}
arrests_kc |>
  mutate(
    departure_country=str_to_title(departure_country),
    citizenship_country=str_to_title(citizenship_country)
  ) |>
  filter(!is.na(departure_country) & departure_country!=citizenship_country) |>
  filter(post_0125) |>
  count(citizenship_country,departure_country,name="number_of_people") |>
  arrange(desc(number_of_people)) |>
  make_output_cols() |>
  gt() |>
  tab_header("Arrests Leading to Departures to Third Countries (Feb 2025-Feb 2026)")

```

#### Age and gender

Each column in this table shows the percent each age group makes up of all
arrests within a gender. For example, 6-10 year olds make up 4.5% of all female
arrests.

For both men and women, people in their 20s and 30s make up over half of all
arrests.
```{r,collapse=TRUE}
arrests_kc |>
  count(age_bin,gender) |>
  pivot_wider(names_from=gender,values_from=n) |>
  column_percents() |>
  make_output_cols() |>
  select(-`Unknown`) |>
  gt() |>
  tab_header("Percent of Arrests by Age (Oct 2022-Feb 2026)") |>
  sub_values(
    fn = function(x) is.na(x),
    replacement = "-"
  ) |>  
  fmt_percent(decimals = 1,scale_values = FALSE)


```


The below table and chart show the change in the number of arrests by age. 
Arrests of people under 18 have decreased since the end of January 2025. 
However, apprehensions of 18-21 year olds have more than tripled, as have 
arrests of those 50 and older.

```{r,collapse=TRUE}
age_tbl <- arrests_kc |>
  comp_table(age_bin,sort_by_var = TRUE,str_title = FALSE) |>
  make_output_cols() 

age_tbl |>
  add_before_after_change() |>
  gt_defaults("Average Monthly Arrests by Age")
```



```{r}
age_tbl |>
  pivot_for_plot() |>
  ggplot(aes(x=`Age Bin`,y=`value`,label=signif(`value`,2),fill=`Period`)) +
  geom_col(position="dodge",width=.7) +
  scale_fill_manual(values=color_vec) +
  geom_text(position = position_dodge(width = .8),vjust=-.6,size=2.75) +   
  labs(title="Average Monthly Arrests by Age",
       subtitle=paste0(
         "Kansas City Area, ",
         format(first_date,"%b %Y"),"-",
         format(last_date,"%b %Y")
        ),
       x="Age Bin",y="Average Monthly Arrests") 

```


Looking at the number of arrests by gender shows that both men and women are being 
arrested at much higher rates, with a larger increase in arrests of men.
```{r,collapse=TRUE}
gender_tbl <- arrests_kc |>
  comp_table(gender,sort_by_var = TRUE) |>
  make_output_cols() 

gender_tbl |>
  add_before_after_change() |>
  gt_defaults("Average Monthly Arrests by Gender")

```

```{r}
gender_tbl |>
  pivot_for_plot() |>
  ggplot(aes(x=`Gender`,y=`value`,label=signif(`value`,2),fill=`Period`)) +
  geom_col(position="dodge",width=.7) +
  scale_fill_manual(values=color_vec) +
  geom_text(position = position_dodge(width = .7),vjust=-.6,size=2.75) +   
  labs(title="Average Monthly Arrests by Gender",
       subtitle=paste0(
         "Kansas City Area, ",
         format(first_date,"%b %Y"),"-",
         format(last_date,"%b %Y")
        ),
       x="Gender",y="Average Monthly Arrests") 

```


## Apprehension Tactics
### Locations
There are a very large number of values in the `apprehension_method` column.
```{r,collapse=TRUE}
arrests_kc |> 
  count(apprehension_method) |> 
  arrange(desc(n)) |>
  rename(`Apprehension Method`=apprehension_method,`Number of Arrests`=n) |>
  gt()

```


To simplify this, creating groups of apprehension methods to get easier-to-use place information. Per the Deportations Data Project [FAQ](https://deportationdata.org/docs/ice.html#faq)

> ...there are two indicators that may be useful: in the arrests table when 
“Apprehension Method” is “Located” or “Non-Custodial Arrest” we think that these 
records are more likely to indicate arrests in the community.


Based on this and other research, setting those values as community arrests, 
[CAP Incarcerations](https://www.americanimmigrationcouncil.org/fact-sheet/criminal-alien-program-cap-immigration-enforcement-prisons-and-jails/) and custodial arrests as prison/jail arrests and creating an "Other" category for the rest.

```{r,collapse=TRUE}
arrests_kc <- arrests_kc |>
  mutate(
    apprehension_location = factor(
      case_when(
        apprehension_method %in% c("LOCATED","NON-CUSTODIAL ARREST") ~ 
          "Community",
        str_detect(apprehension_method,"^CAP .* INCARCERATION$") ~ 
        "Prison/Jail",
        apprehension_method == "CUSTODIAL ARREST" ~ 
          "Prison/Jail",      
        TRUE ~ "Others"
      ), levels = c("Community","Prison/Jail","Others")
    )
  )
  
arrests_kc |>
  count(apprehension_location)

```


Both community and prison/jail arrests have seen large increases since Jan 2025, 
with the former occurring twice as often each month and the latter more than tripling.

```{r,collapse=TRUE}

arrests_kc |>
  make_output_cols() |>
  ggplot(aes(x=`Apprehension Month`,fill=`Apprehension Location`)) +
  geom_bar(stat="count") +
  scale_fill_manual(values=loc_color_vec) +  
  inaug_line + 
  add_inaug_text(as.Date("2024-06-03")) +
  labs(title="Average Monthly Arrests by Location Type",
       subtitle=paste0(
         "Kansas City Area, ",
         format(first_date,"%b %Y"),"-",
         format(last_date,"%b %Y")
        ),
       y="Number of Apprehensions") +
  scale_x_date(date_breaks = "6 months", date_labels =  "%b %y") +
  ylim(0,NA)


```

```{r,collapse=TRUE}
loc_tbl <- arrests_kc |>
  comp_table(apprehension_location,sort_by_var = TRUE) |>
  make_output_cols() 

loc_tbl |>
  arrange(desc(`After Jan 2025`)) |>
  add_before_after_change() |>
  gt_defaults("Average Monthly Arrests by Location Type")

```



### Daily peaks
Looking at arrests over time we can see that certain months have peaks in arrests,
particularly arrests within the community. Breaking this out by specific days 
shows that June 3rd - June 4th and June 25th, 2025 had the highest number of 
community-based arrests, though August 2025 and February 2026 both have multiple days with at least 5 arrests.

There are a handful of days prior to Feb 2026 with 5 or more arrests, though the 
large majority have occurred during the current administration.


```{r,collapse=TRUE}
arrests_kc |>
  filter(apprehension_location=="Community") |>
  count(apprehension_date) |>
  filter(n>=5) |> 
  arrange(desc(n)) |>
  mutate(apprehension_date=format(apprehension_date,"%B %d, %Y")) |>
  rename(`Apprehension Date`=apprehension_date,
         `Number of Arrests`=n) |>
  gt()

  
```


```{r,eval=FALSE, include = FALSE}
#Reviewing the arrests from June 3rd and 4th, though there don't seem to be any obvious 
#patterns.

arrests_kc |>
  filter(apprehension_location=="Community") |>
  filter(apprehension_date == as.Date("2025-06-03") | 
           apprehension_date == as.Date("2025-06-04")) |>
  select(apprehension_state,apprehension_site_landmark,toa_current_duty_site,
         final_program,apprehension_method,apprehension_criminality,
         birth_year,age_bin,citizenship_country,gender) |>
  make_output_cols() 
```


### Daily and hourly trends

Creating hourly bins for easier plotting.
```{r,collapse=TRUE}

hr_bins <- seq(0,23,2)
hr_labs <- c(
  "12 AM - 2 AM",
  "2 AM - 4 AM",
  "4 AM - 6 AM",
  "6 AM - 8 AM",
  "8 AM - 10 AM",
  "10 AM - 12 PM",
  "12 PM - 2 PM",
  "2 PM - 4 PM",
  "4 PM - 6 PM",
  "6 PM - 8 PM",
  "10 PM - 12 AM"
  )
arrests_kc <- arrests_kc |>
  mutate(
    apprehension_dow=wday(apprehension_date_time,label=TRUE),
    apprehension_hour=hour(apprehension_date_time),
    apprehension_block=cut(
      apprehension_hour,
      breaks=hr_bins,
      labels=hr_labs,
      right=TRUE,
      include.lowest=TRUE,
      dig.lab=0,
      ordered_result=TRUE
    )
  )


```


The below chart focuses on arrests during the current administration. 
While there have been arrests every day of the week and at most times throughout
the day, there are clear patterns in ICE activity within communities. Most
arrests occurring on weekday mornings, with particular peaks on Tuesday and
Wednesday from 8-10 AM.

Note that this chart provides a description of what arrest patterns _have_ been,
and is not designed to predict the patterns that they _will_ follow in the 
future.

```{r,collapse=TRUE}
arrests_kc |>
  filter(apprehension_location=="Community") |>
  filter(post_0125) |>
  count(post_0125,apprehension_dow,apprehension_block) |>
  ggplot(aes(x=apprehension_dow,y=apprehension_block)) +
    geom_tile(aes(fill=n)) +
    geom_text(aes(label=n),size=2.75) +
    scale_fill_continuous(low="grey90",high="red") +
    labs(title="Number of Community Arrests by Day and Time",
        subtitle=paste0(
         "Community arrests within the Kansas City area, ",
         format(first_date_admin,"%b %Y"),"-",
         format(last_date,"%b %Y")
        ),
         x="Day of Week",y="Hour",
         fill="Apprehensions") 

```



## Criminality

### Overall criminality
Looking at arrests based on whether people have been charged/convicted of a crime. 
For this section, note that "Other Immigration Violator" indicates the person does not have any charges or convictions against them, per 
the [Deportation Data Project Codebook](https://deportationdata.org/docs/ice/codebook.html).

While all groups have seen increases in arrests, the number of arrests of people
who have been charged with a crime but not yet convicted has skyrocketed.
```{r,collapse=TRUE}
crim_tbl <- arrests_kc |>
  comp_table(apprehension_criminality,sort_by_var = TRUE) |>
  make_output_cols() 

crim_tbl |>
  add_before_after_change() |>
  gt_defaults("Average Monthly Arrests by Criminality")

```


```{r}
crim_tbl |>
  pivot_for_plot() |>
  ggplot(aes(x=`Apprehension Criminality`,y=`value`,
             label=signif(`value`,2),fill=`Period`)) +
  geom_col(position="dodge",width=.7) +
  scale_fill_manual(values=color_vec) +
  geom_text(position = position_dodge(width = .7),vjust=-.6,size=2.75) +   
  labs(title="Average Monthly Arrests by Criminality",
       subtitle=paste0(
         "Kansas City Area, ",
         format(first_date,"%b %Y"),"-",
         format(last_date,"%b %Y")
        ),
       x="Criminality",y="Average Monthly Arrests") 

```


Looking at this by month shows an immediate increase in the number of individuals
arrested with pending criminal charges, the smallest group prior to the current 
administration.

This also shows some clear peaks in arrests of those with neither a criminal charge or conviction, similar to the peaks in community-based arrests shown in the previous section.
```{r,collapse=TRUE}
crim_color_vec <- c("1 Convicted Criminal"="lightcoral",
                    "2 Pending Criminal Charges"="lightblue4",
                    "3 Other Immigration Violator"="darkmagenta")

arrests_kc |>
  mutate(apprehension_criminality=str_to_title(apprehension_criminality)) |>
  make_output_cols() |>
  ggplot(aes(x=`Apprehension Month`,fill=`Apprehension Criminality`)) +
  geom_bar(stat="count") +
  scale_fill_manual(values=crim_color_vec) +  
  inaug_line + 
  add_inaug_text(as.Date("2024-05-20")) +
  labs(title="Number of Apprehensions by Criminality",
       subtitle=paste0(
         "Kansas City Area, ",
         format(first_date,"%b %Y"),"-",
         format(last_date,"%b %Y")
        ),
       y="Number of Apprehensions") +
  scale_x_date(date_breaks = "6 months", date_labels =  "%b %y") +
  ylim(0,NA)

```


### Severity of crimes
For those convicted of a crime, also looking at the severity of that crime, using 
the `case_threat_level` variable. 

Per the Deportation Data Project, this data is potentially based on criteria from a [2011 ICE Memo](https://www.ice.gov/doclib/news/releases/2011/110302washingtondc.pdf)

* Level 1: Aggravated felony
* Level 2: Felony or at least 3 misdemeanors
* Level 3: Misdemeanors


Creating a single variable to combine criminality and severity. 
```{r,collapse=TRUE}
crim_levels <- 5:0
crim_labels <- c(
  "No Criminal Charges",
  "Criminal Charges, No Convictions",
  "Conviction: 1-2 Misdemeanors",
  "Conviction: Felony/3 or More Misdemeanors",    
  "Conviction: Aggravated Felony",
  "Conviction: Unknown Severity"
)

arrests_kc <- arrests_kc |> 
  mutate(crim_summ=
    case_when(
      is.na(apprehension_criminality) ~ NA,
      apprehension_criminality == "3 OTHER IMMIGRATION VIOLATOR" ~ 5,
      apprehension_criminality == "2 PENDING CRIMINAL CHARGES" ~ 4,
      case_threat_level==3 ~ 3,
      case_threat_level==2 ~ 2,    
      case_threat_level==1 ~ 1,
      TRUE ~ 0
    )
  )
arrests_kc <- arrests_kc |>
  mutate(crim_summ=factor(crim_summ,levels = crim_levels,labels=crim_labels))

```



Breaking down arrests by threat level shows that those convicted of only 1-2 
misdemeanors have seen much larger increases in arrests than those convicted of 
more serious crimes. Meanwhile, those who are convicted of an aggravated felony 
actually have the _smallest_ increase in number of arrests out of any group.

```{r,collapse=TRUE}
crim_levels <- 5:0
crim_labels <- c(
  "No Criminal Charges",
  "Criminal Charges, No Convictions",
  "Conviction: 1-2 Misdemeanors",
  "Conviction: Felony/3 or More Misdemeanors",    
  "Conviction: Aggravated Felony",
  "Conviction: Unknown Severity"
)

sev_tbl <- arrests_kc |>
  mutate(crim_summ=
    case_when(
      is.na(apprehension_criminality) ~ NA,
      apprehension_criminality == "3 OTHER IMMIGRATION VIOLATOR" ~ 5,
      apprehension_criminality == "2 PENDING CRIMINAL CHARGES" ~ 4,
      case_threat_level==3 ~ 3,
      case_threat_level==2 ~ 2,    
      case_threat_level==1 ~ 1,
      TRUE ~ 0
    )
  ) |>
  comp_table(crim_summ) |>
  mutate(crim_summ=factor(crim_summ,levels=crim_levels,
                          labels=crim_labels)) |>
  arrange(crim_summ) |>
  filter(crim_summ!="Conviction: Unknown Severity") |>
  make_output_cols() 

sev_tbl |>
  add_before_after_change() |>
  rename(`Criminality`=`Crim Summ`) |>
  gt_defaults("Average Monthly Arrests by Criminality/Severity") 

```

```{r}
sev_tbl |>
  pivot_for_plot() |>
  ggplot(aes(x=`Crim Summ`,y=`value`,
             label=signif(`value`,2),fill=`Period`)) +
  geom_col(position="dodge",width=.7) +
  scale_fill_manual(values=color_vec) +
  geom_text(position = position_dodge(width = .7),vjust=-.6,size=2.75) +   
  scale_x_discrete(labels = scales::wrap_format(10)) + 
  labs(title="Average Monthly Arrests by Criminality/Severity",
       subtitle=paste0(
         "Kansas City Area, ",
         format(first_date,"%b %Y"),"-",
         format(last_date,"%b %Y")
        ),
       x="Criminality/Severity",y="Average Monthly Arrests") 
```


