---
title: "KC Analysis"
output: html_document
date: "2026-04-04"
---

# Analysis of ICE data in Kansas City


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

High-level Plan
1. Start with apprehension data and filter down based on toa_curent_duty_site
  should compare with other indicators
2. Left join on detention data
3. Look at time/distance from apprehension to first detention location
4. Start with detention data, and see who is booked into KC area facilities
5. Check the overlap of this group with #2 
(e.g. those with a first stay in KC but an apprehension elsewhere and vice versa)
  See if documentation discusses why this could happen
6. Replicate #3
7. Perform other analyses flagged in scripts

TODO: For Arrests, look at toa_current_duty_site (think this is a new column in March 26 data)
TODO: Look at individuals in one dataset but not others, especially if it seems
like they should be (eg everyone should have an apprehension)
TODO: Keep ppl who have ever been apprehended in KC in addition to apprehensions 
that occur in KC


## Set up environment
```{r,echo=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(arrow)
source("fns.R")
```


## Get Data
TO DO: Get this from arrests_kc output from other script
```{r}
arrests <- read_parquet(here("data/cleaned","arrests.parquet"))

```


## Detentions 
NOTE: Keeping a subset of columns, may want to add more at a later date.
```{r}
detention_orig <- read_parquet(here("data/march-2026",
                                    "detention-stays-latest.parquet"),
col_select = c(unique_identifier,n_stints,detention_facility_codes_all,
stay_book_in_date_time,stay_book_out_date_time,entry_status,
offense_ina_236c_yes_no,case_ina_236c_yes_no,msc_sentence_days,felon,
bond_posted_date,bond_posted_amount,detainee_classification,final_charge,
book_in_criminality,entry_date,book_in_site,
detention_facility_code_first,detention_facility_first,
book_in_date_time_first,book_in_date_time_last,book_in_aor,n_stays)
) |>
  clean_names() |>
  clean_strings()

```
TO DO: add basic cleaning (or do that in another script and read in cleaned dataset)



### Join Detentions to Arrests

Check Uniqueness
```{r}
arrests_kc |> count(unique_identifier) |> filter(n>1)
detention_orig |> count(unique_identifier) |> filter(n>1)
```

Since both arrests and detentions have multiple IDs, need to avoid a m:m join.


#### Prepare for Joins
Start by separating duplicated from joining unique arrests to unique detentions

```{r}
arrests_kc <- arrests_kc |> mutate(n_appr_kc=n(),
                                   .by=unique_identifier)

arrests_kc_ind <- arrests_kc |> filter(n_appr_kc==1)
arrests_kc_mult <- arrests_kc |> filter(n_appr_kc>1)
detention_ind <- detention_orig |> filter(n_stays==1)
detention_mult <- detention_orig |> filter(n_stays>1)
```


```{r}
nrow(arrests_kc)
nrow(arrests_kc_ind)
nrow(arrests_kc_mult)
```

Convenience functions for matching
TO DO: Add a function to automatically confirm row counts match expected
```{r}
find_correct_detention <- function(arr_det_df,by){
  arr_det_df <- arr_det_df |>
  mutate(appr_det_diff=difftime(book_in_date_time_first,apprehension_date_time,
                                units = "hours")) |>
  filter(!appr_det_diff<(-24*7)) |>
  filter(row_number()==1,.by={{by}}) |>
  select(-appr_det_diff)
}

remove_matched <- function(df,found_vec){
  df |> filter(!unique_identifier %in% found_vec)
}


```


#### Unique Arrests
First join those with a 1:1 relationship
```{r}
arr_det_ind <- inner_join(arrests_kc_ind,detention_ind,
                          by="unique_identifier",relationship="one-to-one")
arrests_kc_ind <- remove_matched(arrests_kc_ind,arr_det_ind$unique_identifier)  
```

```{r}
nrow(arrests_kc_ind)
nrow(arr_det_ind)
```

Now join unique arrests to non-unique detentions and keep the closest that isn't
more than a week before the arrest.


```{r}
arr_det_ind_2 <- inner_join(arrests_kc_ind,detention_mult,
                            by="unique_identifier",relationship="one-to-many")
```


```{r}
arr_det_ind_2 <- find_correct_detention(arr_det_ind_2,unique_identifier)
arrests_kc_ind <- remove_matched(arrests_kc_ind,arr_det_ind_2$unique_identifier)  

```


```{r}
nrow(arr_det_ind_2)
nrow(arrests_kc_ind)

```
Check if the remaining IDs in arrests_kc_ind appear in detention at all (they shouldn't)
```{r}
remaining_ids_ind <- arrests_kc_ind$unique_identifier
remaining_ids_ind[remaining_ids_ind %in% detention_orig$unique_identifier]
```
Combine datasets
```{r}
arrests_kc_ind <- bind_rows(arrests_kc_ind,arr_det_ind,arr_det_ind_2)
```

```{r}
rm(arr_det_ind,arr_det_ind_2)
```


#### Multiple Arrests

##### One Detention
Join multiple arrests to individual detentions
```{r}
arrests_kc_mult <- arrests_kc_mult |> 
  mutate(appr_id = paste(unique_identifier,appr_ind,sep="_"))

arr_det_mult <- inner_join(arrests_kc_mult,detention_ind,
                           by="unique_identifier",relationship="many-to-one")

```


TO DO: Add this to fn to remove found IDs
```{r}
arr_det_mult_check <- find_correct_detention(arr_det_mult,unique_identifier)

found_vec <- arr_det_mult_check$appr_id
arrests_kc_mult <- arrests_kc_mult |> 
  filter(!appr_id %in% found_vec) 

```


##### Multiple Detentions
Join multiple arrests to multiple detentions, by rounded month of arrest.

Remove very small number of records where this still yields multiple arrests.
```{r}

### add rounded event period
unit <- "month"
arrests_kc_mult <- arrests_kc_mult |>
  mutate(event_rounded=round_date(apprehension_date_time,unit=unit))

detention_mult <- detention_mult |>
  mutate(event_rounded=round_date(book_in_date_time_first,unit=unit))

### identify multiples
has_mult_id_appr_rd <- arrests_kc_mult |>
  count(unique_identifier,event_rounded) |>
  filter(n>1) |>
  select(unique_identifier) |>
  unlist()

### separate
arrests_kc_mult_appr_rd <- arrests_kc_mult |>
  filter(unique_identifier %in% has_mult_id_appr_rd)

arrests_kc_mult <- arrests_kc_mult |>
  filter(!unique_identifier %in% has_mult_id_appr_rd)

```

Perform join

NOTE: Not sure this will always be 1-1. Need to consider it more.
```{r}
arr_det_mult2 <- inner_join(arrests_kc_mult,detention_mult,
                      by=c("unique_identifier","event_rounded"),
                      relationship="one-to-one")

```

Review matches
```{r}
arr_det_mult2 |>
  select(unique_identifier,
         apprehension_date_time,book_in_date_time_first) |>
  arrange(unique_identifier,
         apprehension_date_time,book_in_date_time_first)  
```

Identify still-unmatched arrests
```{r}
matched_mult <- arr_det_mult2$unique_identifier
arrests_kc_mult <- arrests_kc_mult |>
  filter(!unique_identifier %in% matched_mult)
```



#### Combine
```{r}
arrests_kc <- bind_rows(
  arrests_kc_ind,
  arrests_kc_mult,
  arr_det_mult,
  arr_det_mult2
)

# remove unneeded data
rm(
  arrests_kc_ind,
  arrests_kc_mult,
  detention_ind,
  detention_mult,
  arr_det_mult,
  arr_det_mult2,
  arr_det_mult_check,
  arrests_kc_mult_appr_rd
)
```

# Analysis 
## Arrest-to-Detention Analysis
Time from arrest to detention
```{r}
arrests_kc <- arrests_kc |>
  mutate(
    det_appr_hrs=as.numeric(difftime(
      book_in_date_time_first,
      apprehension_date_time,
      units="hours")),
    det_appr_days=as.numeric(difftime(
      book_in_date_time_first,
      apprehension_date_time,
      units="days"))    
  ) 
```


```{r}
summary(arrests_kc$det_appr_hrs)

arrests_kc |>
  filter(det_appr_hrs>=-1 & det_appr_hrs<=1000) |>
  ggplot(aes(x=det_appr_hrs)) +
  geom_histogram()

```

```{r}
arrests_kc |>
  filter(det_appr_days>=1) |>
  select(unique_identifier,apprehension_date_time,book_in_date_time_first,
         apprehension_site_landmark,detention_facility_first
        )
```


## Identify Detention Locations of KC Arrests
```{r}
kc_inds <- arrests_kc$unique_identifier
detention_kc <- detention_orig |>
  filter(unique_identifier %in% kc_inds)

detention_kc |>
  count(book_in_site) |>
  arrange(desc(n))

detention_kc |>
  count(det) |>
  arrange(desc(n))


```




# outcomes (in comparison to country?)

COLUMNS NEEDED
* LENGTH OF STAY 

OTHER ANALYSES
* Merge detainers and detention to see how many requests were approved by local law enforcement
